---
title: "Streamlining Design and Maximizing Success for Agile Test and Learn"
author: "Ncpeifer"
date: "03/13/2019"
output: html_document
---

## Abstract

Test and learn analysis is used widely throughout industry as a way to quantitatively understand the preferences of a retailer’s customer base. Successful testing requires accurate assumptions (response rates, anticipated difference between variants) along with finely tuned execution parameters (audience size, test / control split, test iterations and timing). The test design phase is arguably the most critical and requires timely collaboration between marketing strategists, statisticians, data analysts and operations experts to agree on the right parameters for a successful test execution. Through our work with large retailers, we have come up with an approach to significantly accelerate the test and learn design phase, using a set of functions developed using R statistical software. Our approach increases the speed and ease of effect size calculations across any number of scenarios, allowing marketing strategists to quickly rule out doomed test designs and hone in on the right parameters to maximize the chances of a successful test.

## Introduction

Formally speaking, A/B testing is used compare two or more versions of a single variable. A recent engagement with a large retailer lead our team to employ A/B testing to compare the effectiveness of different email campaigns at bringing existing customers back for follow-up purchases. The goal of the group is to test quickly and strategically, making each campaign as personalized as possible to it's core audience. For example, we may want to target Men's shoppers who bought a suit in the last 3 months and compare an offer for tailoring services against a discount on dress shoes. We need to put this test into market quickly because the holiday season is fast approaching, and we have a backlog of additional test ideas lined up as soon as this one is done. 

To get this done, we assembled an agile team of marketing strategists, operations experts and data scientists. While the marketing strategists are ideating on creative content and the desired target audience, the objective of the data scientist is to assess whether or not the anticipated design has a good chance of yielding a detectible difference between A and B. If the current design is unlikely to yield a result, the whole team needs to quickly pivot to a different design and the assessment will begin again. 

## Problem

A number of factors make this test design phase a complicated task for the data scientist:

**1.** Each test is designed for customers who exhibit very specific behaviors, so we are constantly dealing with **small audience sizes**

**2.** With highly personalized tests, there is a **large variability in baseline response rates and the anticipated lift** of each test

**3.** Some tests carry additional costs of implementation (e.g. promotions or gift cards) which means we need to **vary the test and control group split**

**4.** The agile framework is designed to launch three tests per week, meaning **rapid test design sprints** 

Many great resources are available for determining the audience size or effect size needed given specific input values (for example: https://www.evanmiller.org/ab-testing/sample-size.html). But plugging in 30 combinations of inputs into these calculators and comparing results is cumbersome at best and severly hinders the design process.

## Solution

To help streamline test design, we came up with a way to see how big of a difference in response rate between A and B we would need to see to confidently confirm or reject a test hypothesis **under any number of input scenarios**. 

### Set up

Our goal is to take varying design inputs into consideration and return a range of possible execution options for the team to choose from during ideation. The main inputs we needed to incorporate into the assessment are: 

**1. Audience size** - The number of customers who qualify for the test

**2. Test proportion** - The split between Test and Control customers (Test receives A content, control receives B)

**3. Baseline response** - The expected rate at which customers normally come back in and shop, measured as a probability between 0 and 1

The resulting assessment should quantify the difference between A and B needed in order to confirm or reject a test hypothesis. 

### Execution

Say we want to promote a new tailoring service in our store. We define a general hypothesis that a 15% off promotion for services will cause more customers to use the tailoring service than a purely informational email promoting the service. The test group will get the 15% off promotion while the control will get a standard email informing them about tailoring. 

#### Define the inputs

##### Audience Size

Our ideal target audience is the 10,000 customers who bought a suit as their only purchase in the past year. If needed, we could loosen our criteria to include the additional 10,000 customers who bought a suit and shoes in the past year, or even the additional 55,000 customers who bought dress shirts but did not buy a suit. So for our assessment, we'll look at audience sizes ranging from 10k to 75k. 

##### Test Proportion

The test campaign includes a 15% off promotion which decreases profit margin, so we favor an unbalanced test to minimize the number of customers receiving the test promotional content as much as possible. For our assessment we'll look at a full range of splits: 10/90, 20/80, 30/70, 40/60, and 50/50. 

##### Baseline Response

Finally, we know the general baseline response rate is 12% for existing shoppers to use similar services. While we know that promotions generally increase response rates by 1-3 percentage points, we want to be conservative with this test because the tailoring service is new, so we'll leave this constant at 12%. 

#### Define the functions

##### Effect size calculation

The first function uses a two proportion, unbalanced calculation from the [pwr] package in R to return the effect size needed to detect statistical significance. We use a two proportion test because our response variable is a probability between 0 and 1. We use an unbalanced sample because we want to see different test and control sample sizes (i.e. not limited to a 50/50 split). Note that we're using a power of 80%. Because the cost of sending an email is negligible, we can afford to have more wiggle room for error than the standard 90% or 95% thresholds. 

[Note that this approach accommadates variability in the primary effect size parameters of audience size, test proportion, and baseline response rate and is built for two-proportion, unbalanced tests. Other test types, such as comparison of a continous outcome, have not been explored here, but could be accommodated within the same framework.]

```{r setup, message=FALSE}
library(plotly)
library(ggplot2)
library(pwr)
library(tidyverse)
library(kableExtra)

unbalanced <- function(x,y) {
  pwr.2p2n.test(h = NULL, 
                n1 = x*y,
                n2 = x*(1-y),
                sig.level = 0.2, 
                power = .80, 
                alternative = "greater")
}
```

The second function solves the effect size formula for the required test response rate needed to confirm or reject our hypothesis. We use *Cohen's h* formula for effect size since we're comparing differences between two proportions (https://en.wikipedia.org/wiki/Cohen%27s_h). 

``` {r}
# Effect size formula: h = 2*asin(sqrt(p2))-2*asin(sqrt(p1))
# Solving for test response rate (p2) gives: sin((h + 2*asin(sqrt(p1)))/2)^2
p2.response <- function(h,p1) {
  sin((h + 2*asin(sqrt(p1)))/2)^2
}
```

#### Set the input parameters

```{r}
# Adjust these based on test design
audience_size <- c(10000,20000,75000)
test_proportion <- c(0.1,0.2,0.3,0.4,0.5)
baseline_rate = c(0.12)
```

#### Create a data set with all possible test designs

```{r}
setup <- tibble(audience_size) %>%
    merge(tibble(test_proportion)) %>%
    merge(tibble(baseline_rate))
```

#### Apply the effect size calculations to each row of the dataset

```{r}
Sys.setenv("plotly_username"="ncpeifer")
Sys.setenv("plotly_api_key"="ooSWEvq5vnknc1Br8TF7")

calculations <- setup %>%
    mutate(h_power = pmap(list(audience_size, test_proportion), unbalanced)) %>%
    mutate(effect_size = unlist(map(h_power,"h")),
           test_n = unlist(map(h_power, "n1")),
           control_n = unlist(map(h_power, "n2")),
           total_n = test_n + control_n,
           test_rate = unlist(map2(effect_size,baseline_rate,p2.response)),
           lift_needed = (test_rate - baseline_rate) / baseline_rate) %>%
    select(-h_power)

calculations %>%
  kable() %>%
  kable_styling(bootstrap_options = "striped", full_width = F)

api_create(calculations, filename = "r-docs-ab-lift-data")
```

Now we have a clean table showing the test audience size (test_n), along with the percentage lift over the baseline response rate needed in order to confirm or reject the hypothesis. This is technically all the info we need, but no one likes to read a boring old table...

#### Create an interactive plot for all possible designs

Using [ggplot] and [plotly] visualization packages, we can create a nice illustration of all the test options with interactivity to let us hover over different options to see all the pertinent detials. 

```{r}
lift_plot <- ggplot(data = calculations %>%
           mutate(text = paste("Audience Size: ", audience_size, "\nTest Proportion: ",  test_proportion,  
                               "\nTest Audience: ", test_n, "\nControl Audience: ",  control_n, 
                               "\nBaseline Response: ", baseline_rate, "\nResponse Rate Needed: ", round(test_rate,4),
                               "\nLift Needed: ", round(lift_needed,4), sep="")), 
         aes(x = as.factor(audience_size), text = text)) + 
    geom_line(mapping = aes(y = lift_needed, group = as.factor(test_proportion), color = as.factor(test_proportion))) + 
    geom_point(aes(y = lift_needed, group = as.factor(test_proportion), color = as.factor(test_proportion))) +
    xlab("Total Audience Size") + 
    ylab("Lift Needed") +
    labs(color = "Test Proportion")

plotly_lift_plot <- ggplotly(lift_plot, tooltip="text")

plotly_lift_plot

api_create(plotly_lift_plot, filename = "r-docs-ab-lift-plot")

import chart_studio.tools as tls
tls.get_embed('https://plot.ly/~elizabethts/9/') #change to your url

```

## Results

Looking at the resulting plot, we can quickly see the pros and cons of each design option. Using the 10,000 suit-only shoppers, our campaign would need to drive at least a 9% lift with a 50/50 test and control split. That type of lift is unlikely given our knowledge of past campaigns, and because of the promotional offer we would much rather limit the test recipients to less than 50% of the audience. Because of that, we might suggest relaxing the audience criteria to include the additional 65,000 suit + shoes and dress shirt shoppers to reach a 75,000 audience size with a 20/80 test and control split, where we need a much more reasonable 5% lift to confirm or reject the hypothesis. 

## Let's go a bit further

We have just looked at an example using specific audience sizes based on three initial sets of criteria. But what if we go a step further and see how the required lift changes across a broad range of audience sizes and varying baseline rates? We can adjust the inputs and see. Note for this exercise we'll hold the test proportion constant at a 50/50 split.

```{r}
audience_size <- c(seq(from = 1000, to = 100000, by = 5000))
test_proportion <- c(0.5)
baseline_rate = c(0.05, 0.1, 0.15)

setup2 <- tibble(audience_size) %>%
    merge(tibble(test_proportion)) %>%
    merge(tibble(baseline_rate))

calculations2 <- setup2 %>%
    mutate(h_power = pmap(list(audience_size, test_proportion), unbalanced)) %>%
    mutate(effect_size = unlist(map(h_power,"h")),
           test_n = unlist(map(h_power, "n1")),
           control_n = unlist(map(h_power, "n2")),
           total_n = test_n + control_n,
           test_rate = unlist(map2(effect_size,baseline_rate,p2.response)),
           lift_needed = (test_rate - baseline_rate) / baseline_rate) %>%
    select(-h_power)
options(scipen = 99)
lift_plot2 <- ggplot(data = calculations2 %>%
           mutate(text = paste("Audience Size: ", audience_size, "\nTest Proportion: ",  test_proportion,  
                               "\nTest Audience: ", test_n, "\nControl Audience: ",  control_n, 
                               "\nBaseline Response: ", baseline_rate, "\nResponse Rate Needed: ", round(test_rate,4),
                               "\nLift Needed: ", round(lift_needed,4), sep="")), 
         aes(x = audience_size, text = text)) + 
    geom_line(mapping = aes(y = lift_needed, group = as.factor(baseline_rate), color = as.factor(baseline_rate))) + 
    geom_point(aes(y = lift_needed, group = as.factor(baseline_rate), color = as.factor(baseline_rate))) +
    scale_x_continuous(name = "Total Audience Size", breaks=seq(0,100000,10000)) +
    ylab("Lift Needed") +
    labs(color = "Baseline Rates")
plotly_lift_plot2 <- ggplotly(lift_plot2, tooltip="text")
plotly_lift_plot2

api_create(plotly_lift_plot2, filename = "r-docs-ab-lift-plot-2")
```

### Interpretation

As expected, the smaller the baseline rate, the higher the amount of lift you would need to see to detect a difference. What's interesting for the marketing strategist is the non-linear relationship between audience size and lift required. This relationship is important to consider when designing extremely personalized campaigns targeting small audiences. For example, with a baseline response rate of 10% and an audience of 1,000 customers, you would need to see at least a 34% lift to confirm your hypothesis. However, increasing that audience by 5,000 customers reduces the lift needed to 13%. With the same 10% baseline and 20,000 customers in your audience, you would need a 7% lift and the benefit of an additional 5,000 customers is a relatively small difference of 0.7 percentage points. 

## Conclusion

We have developed an approach for A/B test design that accelerates sample and effect size calculations across any number of design scenarios and provides an interactive visual of all possible designs. This approach greatly increased the speed at which our test and learn team can *rule out doomed test designs* and hone in on the right parameters to *maximize the chances of a successful test*. 


